---
title: "HW2 TJ Jiang"
output:
 pdf_document:
  latex_engine: xelatex
---

Load necessary libraries

```{r}
library(ISLR)
library(car)
```

## Problem 1

Chapter 4, Exercise 4 (p. 168).

### a) 
For the cases where 0.5 < X < 0.95, the average will be 10%. Otherwise, we form an integral as such: 

$$\int_{0}^{0.05} 100x + 5 dx$$
Which equals 0.375, multiplied by 2 for two intervals: when x < 0.05 and when x > 0.95
Thus on average our prediction is (0.1 * 0.9 + 0.00375 * 2) * 100 = 9.75%

### b) 
0.0975^2 * 100 = 0.95%

### c) 
0.0975^100 * 100 = (7.95 e-100)%

### d)
Our results show that as dimensionality increases, the number of datapoints that are close in all dimensions to the response variable decreases exponentially. 

### e)
The length will be 0.1^(1/p)
e.g. for p=1, l = 0.1^(1) = 0.1
for p=100, l = 0.1^(1/100)

Problem 2

Chapter 4, Exercise 6 (p. 170).

### a)
$$
\frac{e^{B_0+B1X_1+...+B_pX_p}}{1+e^{B_0+B_1X_1+...+B_pX_p}}\\
\frac{e^{-6+0.05*40+3.5}}{1+e^{-6+0.05*40+3.5}}\\
=0.3775
$$

### b)
$$
0.5 = \frac{e^{-6+0.05*X_1+3.5}}{1+e^{-6+0.05*X_1+3.5}}\\
0.5*({1+e^{0.05*X_1-2.5}}) = e^{0.05*X_1-2.5}\\
0.5 = 0.5e^{0.05*X_1-2.5}\\
log(1) = log(e^{0.05*X_1-2.5})\\
0 = 0.05*X_1 - 2.5\\
X_1 = 2.5/0.05 = 50 hours
$$

The student in part a) needs to study 50 hours to have a 50% chance of earning an A

## Problem 3

Chapter 4, Exercise 8 (p. 170).

With K=1, the decision boundary is highly flexible. In general, as flexibility increases, the training error will decline but the testing error will increase. A KNN classifier with K=1 has a training error of 0, as every observation will simply cluster with itself. This means that the test error was 36%. 

**Thus, I would choose the 30% test error logistic regression classifier.**


## Problem 4
Updated: only a, b, c, d are required
Chapter 4, Exercise 10 (p. 171). In part (i), please be concise; only describe and provide the output of your best prediction.

### a)

Quick look at correlation matrix

```{r, fig.width=10, fig.height=10}
class(Weekly)
scatterplotMatrix(Weekly, main="Weekly Dataset Scatterplot Matrix")
```



Problem 5

Chapter 5, Exercise 5 (p. 198).



Problem 6

Chapter 5, Exercise 6 (p. 199)



Problem 7

Chapter 5, Exercise 8 (p. 200)



Problem 8

Chapter 5, Exercise 9 (p. 201)